AttCRISPR : a spacetime interpretable model for prediction of sgRNA on-target activity

### 2020-11-18 Paper/main.tex
11-18 paper

\abstract{\textbf{Motivation:} More and more Cas9 variants with higher specificities are developed to avoid the off-target effect, which bring a significant volume of experimental data. 
Conventional machine learning performs poorly on these datasets, while the methods based on deep learning are often lack of interpretability, 
which makes researchers have to trade-off accuracy and interpretability. 
It is necessary to develop a method which can not only match deep learning based methods in performance, 
but also with strong interpretability that can be comparable to conventional machine learning methods.\\
\textbf{Results:} To overcome these problems, we design and implement AttCRISPR, an intrinsic interpretable method based on deep learning to predict the on-target activity. 
The advantage of AttCRISPR lies in using the ensemble learning strategy to stack available encoded-based methods and embedded-based methods with strong interpretability. 
Comparison with the state-of-the-art methods (WT-SpCas9, eSpCas9(1.1), SpCas9-HF1), AttCRISPR can achieve an average spearman value of 0.872, 0.867, 0.867, respectively on several public datasets, which is superior to these methods. 
Furthermore, benefits from two attention modules, spatial and temporal attention modules, AttCRISPR has strong interpretability. 
Through them, we can understand the decisions made by AttCRISPR at both global and local levels without other post hoc explanations techniques. 
With the trained model, we reveal the preference for each position-dependent nucleotide on the sgRNA (short guide RNA) sequence in each dataset at global level. 
And at local level, we prove that interpretability of AttCRISPR can be used to guide researcher to design sgRNA with higher activity. \\
\textbf{Availability:} The example code are available at \href{https://github.com/South-Walker/AttCRISPR}{https://github.com/South-Walker/AttCRISPR}\\
\textbf{Contact:} \href{zrjiang@cs.ecnu.edu.cn}{zrjiang@cs.ecnu.edu.cn}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

At global level, an important question we expect AttCRISPR to answer is which nucleotide it prefers at each position on the sequence. 
In fact, \citeauthor{wang2019optimized} has already answered this question in detail with the DeepSHAP method.
Leaving aside nucleotides, just consider the importance of each position, the method of \citeauthor{Liu2019} have gone some way too. 
While our method is not based on the post hoc explanations techniques and input perturbations, the only work we need to do is to get the first-order preference $A$ generated by the attention module.
Specifically, we use the first-order preference $A_s$ generated by the spatial attention module instead of  $\tilde{A}$ generated by the temporal attention module. 
The latter is in a higher dimensional dense space which make it difficult to understand. 
In practice, we input every sgRNA into the spatial AttCRISPR, in order to obtain the $A_s$ from the spatial attention module and take its mean value. 
Then we rescale it through Z-score to obtain a standardization value and the final result is shown in Figure~\ref{fig:spatialattention}.

@@

For the first question, we input the least-active sgRNA in Supplementary Tab. S1 (with the index of 8493 and the activity of 0.831, call source sgRNA for convenience) into the temporal AttCRISPR. 
The score of each position is obtained based on Equation~(\ref{eq:05}) (the calculated symbol with $W$ is Hadamard product instead of dot product, in order to achieve the result in vector form), 
and the results are shown in Figure~\ref{fig:opt}. 
Acording to Figure~\ref{fig:opt}, scores at position 14 and 16 of source sgRNA are significantly below the base line 
(in fact, the scores at position 6 and 11 is also noteworthy, however, we don't find sgRNA in the dataset for comparison). 
If replaced the T at position 14 with C, would generate the same sgRNA as the one with index of 8491, which is with a activity value of 0.861. 
If replaced the T at position 16 with C, would generate the same sgRNA as the one with index of 8492, which is with a activity value of 0.869. 
Therefore, we could conclude that the local interpretability is helpful for us to optimize the sgRNA without exhaustively search.

>>>>

\abstract{\textbf{Motivation:} More and more Cas9 variants with higher specificities are developed to avoid the off-target effect, which bring a significant volume of experimental data. 
Conventional machine learning performs poorly on these datasets, while the methods based on deep learning are often lack of interpretability, 
which makes researchers have to trade-off accuracy and interpretability. 
It is necessary to develop a method which can not only match deep learning based methods in performance, 
but also with strong interpretability that can be comparable to conventional machine learning methods.\\
\textbf{Results:} To overcome these problems, we design and implement AttCRISPR, an intrinsic interpretable method based on deep learning to predict the on-target activity. 
The advantage of AttCRISPR lies in using the ensemble learning strategy to stack available encoded-based methods and embedded-based methods with strong interpretability. 
Comparison with the state-of-the-art methods (WT-SpCas9, eSpCas9(1.1), SpCas9-HF1), AttCRISPR can achieve an average spearman value of 0.872, 0.867, 0.867, respectively on several public datasets, which is superior to these methods. 
Furthermore, benefits from two attention modules, spatial and temporal attention modules, AttCRISPR has strong interpretability. 
Through them, we can understand the decisions made by AttCRISPR at both global and local levels without other post hoc explanations techniques. 
With the trained model, we reveal the preference for each position-dependent nucleotide on the sgRNA (short guide RNA) sequence in each dataset at global level. 
And at local level, we prove that the interpretability of AttCRISPR can be used to guide researcher to design sgRNA with higher activity. \\
\textbf{Availability:} The example code are available at \href{https://github.com/South-Walker/AttCRISPR}{https://github.com/South-Walker/AttCRISPR}\\
\textbf{Contact:} \href{zrjiang@cs.ecnu.edu.cn}{zrjiang@cs.ecnu.edu.cn}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

At global level, an important question we expect AttCRISPR to answer is which nucleotide it prefers at each position on the sequence. 
In fact, \citeauthor{wang2019optimized} has already answered this question in detail with the DeepSHAP method.
Leaving aside nucleotides, just consider the importance of each position, the method of \citeauthor{Liu2019} has gone some way too. 
While our method is not based on the post hoc explanations techniques and input perturbations, the only work we need to do is to get the first-order preference $A$ generated by the attention module.
Specifically, we use the first-order preference $A_s$ generated by the spatial attention module instead of  $\tilde{A}$ generated by the temporal attention module. 
The latter is in a higher dimensional dense space which make it difficult to understand. 
In practice, we input every sgRNA into the spatial AttCRISPR, in order to obtain the $A_s$ from the spatial attention module and take its mean value. 
Then we rescale it through Z-score to obtain a standardization value and the final result is shown in Figure~\ref{fig:spatialattention} and Supplementary Data S3.

@@

For the first question, we input the least-active sgRNA in Supplementary Tab. S1 (with the index of 8493 and the activity of 0.831, call source sgRNA for convenience) into the temporal AttCRISPR. 
The score of each position is obtained based on Equation~(\ref{eq:05}) (the calculated symbol with $W$ is Hadamard product instead of dot product, in order to achieve the result in vector form), 
and the results are shown in Figure~\ref{fig:opt} and Supplementary Data S4, in which scores at position 14 and 16 of source sgRNA are significantly below the base line 
(in fact, the scores at position 6 and 11 is also noteworthy, however, we don't find sgRNA in the dataset for comparison). 
If replaced the T at position 14 with C, would generate the same sgRNA as the one with index of 8491, which is with a activity value of 0.861. 
If replaced the T at position 16 with C, would generate the same sgRNA as the one with index of 8492, which is with a activity value of 0.869. 
Therefore, we could conclude that the local interpretability is helpful for us to optimize the sgRNA without exhaustively search.
